---
title: "Assignmetn 2"
author: "Yineng Chen"
date: "6/13/2020"
output: 
  html_document:
    keep_md: TRUE 
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(data.table)
```

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.The events in the database start in the year 1950 and end in November 2011. 

Details about the data can be found [here][https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf]


In the following analysis, we focus on two queations:

1. Across the United States, which types of events are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

## Load Data

```{r imput}
fileUrl = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

download.file(fileUrl,destfile = "./data/stormdata.csv.bz2")

stormdf = read.csv("./data/stormdata.csv.bz2") %>% janitor::clean_names()

# convert df to dt
stormdt = as.data.table(stormdf)
dim(stormdt)
names(stormdt)
View(stormdt)
str(stormdt)
```


## Data subsetting and processing

We select variables related to population health and economic consequences.

```{r,cache = TRUE}
workdt = stormdt[,.(evtype,fatalities,injuries,
                   propdmg,propdmgexp,cropdmg,cropdmgexp)]

# remove rows with undifined event type and zaro damage
workdt = workdt[(evtype != "?" & 
             (injuries > 0 | fatalities > 0 | propdmg> 0 | cropdmg > 0))]

#  Converting Exponent Columns into Actual Exponents 
cols = c("propdmgexp", "cropdmgexp")
workdt[,  (cols) := c(lapply(.SD, toupper)), .SDcols = cols]

propdmgKey =  c("\"\"" = 10^0,"-" = 10^0, "+" = 10^0,"0" = 10^0,"1" = 10^1,"2" = 10^2,
                 "3" = 10^3, "4" = 10^4,"5" = 10^5,"6" = 10^6,"7" = 10^7,"8" = 10^8,
                 "9" = 10^9,"H" = 10^2,"K" = 10^3,"M" = 10^6,"B" = 10^9)

cropdmgKey =  c("\"\"" = 10^0,"?" = 10^0, "0" = 10^0,
                "K" = 10^3,"M" = 10^6,"B" = 10^9)
  
workdt[,propdmgexp := propdmgKey[as.character(workdt[,propdmgexp])]]
workdt[is.na(propdmgexp), propdmgexp := 10^0 ]

workdt[,cropdmgexp := cropdmgKey[as.character(workdt[,cropdmgexp])]]
workdt[is.na(cropdmgexp), cropdmgexp := 10^0 ]
```

## Most harmful event with respect to population health

### Fatalities

```{r}
hist(workdt[,fatalities])
summary(workdt[,fatalities])

fata <- workdt[order(-fatalities)]

head(fata[,.(evtype,fatalities)],10)
```

